### 2019-05-25 chapter_00_open_science.tex
Final proofread

Collaboration was achieved in the early years of FOSS development through
emailing ``patches'', and centralized Version Control Systems (VCS) near the
turn of the 21\textsuperscript{st} century.  The collective collaboration on
development is today streamlined with the emergence of Distributed Version
Control Systems (DVCS), especially, Git and Mercurial). In web-based source
development platforms like GitHub, Bitbucket and GitLab, DVCS repositories are
enhanced with code-review tools, typically consisting of a issue tracker
and pull-request mechanism.
%
Git has become the \textit{de facto} standard for source management. However,
for the FluidDyn project, we prefer to use Mercurial because we consider it to
be simpler for the beginners and just as powerful for expert users\footnote{See
  \href{https://www.mercurial-scm.org/wiki/GitConcepts}{``a detailed
comparison''} on Mercurial website.}. Simplicity and lowering the learning
barrier for scientists who are not experts in software development is crucial
for our project.
%
Note that, it is now possible to use Mercurial to work with Git repositories,
for example hosted in GitHub and GitLab\footnote{By using the Mercurial
extension \href{http://hg-git.github.io/}{hg-git}.}.
%
Version control tools are designed to work on any kind of text files, not just
codes, so they could even be utilized to write scientific documents with
multiple authors.
%
The use of DVCS in science will lead to long term productivity benefits
\citep{wilson_best_2014}.

@@

Packaging process depends on the language used in a project, however some good
practices are universal, such as licensing, supplying documentation (a README
file at the bare minimum), keeping installation paths flexible and relying on
official build tools \citep{taschuk_ten_2017}.
%
A credible repository to packages as compressed archives or as pre-compiled binary
packages is indispensable to promote code reuse and reproducibility of results.
%
The use of repositories started with the conception of
CPAN and CRAN repositories for the languages Perl and R in 1993 --- based on
the Comprehensive \TeX\ Archive Network (CTAN) model for \TeX\ packages.

Python as a medium for scientific computing owes its success to a rich
ecosystem of third-party packages. \href{https://pypi.org}{Python Package
Index (PyPI)} and \href{https://anaconda.org}{Anaconda Cloud} are major
repositories for delivering and downloading Python packages, assisted by
package managers \pack{pip} and \pack{conda} respectively. The
\fnref{https://packaging.python.org}{``Python Packaging User Guide''} is an
excellent guide to install, package and distribute projects.
% Almost all major
% languages\footnote{\href{https://stackoverflow.com/questions/1693529/%
% list-of-top-repositories-by-programming-language}{https://stackoverflow.com/%
% questions/1693529/list-of-top-repositories-by-programming-language}} have
% similar repositories, with exceptions such as Fortran and Matlab.
Specialized package managers, such as \pack{easybuild} and \pack{spack}, have
also emerged to install scientific libraries and compilers.

\subsubsection{Continuous Integration and documentation} The need for
strong reliability of packages with increasing complexity has forced developers
to use Continuous Integration (CI).
%
When the code depends on a wide variety of external dependencies, CI is useful
to avoid \fnref{https://en.wikipedia.org/wiki/Software\_rot}{software rot},
i.e.\ to avoid obsolescence with newer versions of the compiler or its
dependencies. CI is specifically important for dynamical languages and
open-source code since it allows open-source projects to welcome contributions
from other developers while maintaining a high degree of reliability.
%
Firstly the code is \emph{built} against a known, repeatable emulation
environment on a server. Thereafter, a set of \emph{unit tests} are run for
each commit (modification of the code) to check that no bugs have been
introduced. It is also important to analyse what portion of the source code is
covered by these unit tests --- a method referred to as \emph{code coverage}.
%
Almost all CI platforms are free for open-source projects --- Travis CI,
CircleCI and Appveyor to name a few. Bitbucket and GitLab have built-in CI
implementations. Websites such as Codecov and Coveralls help to chart and
analyse the code coverage history, viewable by anyone. To use these services,
one simply has to write a simple file in YAML format and authenticate the
respective services with ``read'' permissions.

@@

% Mutual aid and knowledge sharing have been important aspects of the web since
% its inception and these practices are now of foremost importance for many
% human activities.
%
%In particular, 
Learning programming and computer science is now greatly aided by websites like
the community driven encyclopaedia \href{https://www.wikipedia.org/}{Wikipedia}
or the community driven forum \href{https://stackoverflow.com/}{Stack
Overflow}, where people write thousands of questions and answers on programming
each week. IRC protocol has continued to sustain the test of time. Instant
messaging channels such as \codeinline{\#python} on Freenode facilitate
interactive discussions and a form of community-level support system --- a role
which used to be filled by mailing lists alone in the past. Alternatives such
as Riot (Matrix protocol), Gitter, and Slack have gained popularity offering
different integrations solutions on top of instant messaging.
%
Through instant messaging a geographically separated team, as in the case of
the FluidDyn project, can achieve near real-time collaboration without having
to arrange video conferences or meetings.

@@

\item \textbf{Code style.} There are official (but, optional) guidelines, named
\fnref{https://www.python.org/dev/peps/pep-0008/}{PEP~8}, advocating for
regularity of the code. Linters and code-formatters can evaluate adherence to
this style.

@@

\item \textbf{Extension support}. It is easy to interface with code written in
several other languages (in particular C, C++ and Fortran).

@@

Thus, Python is one of the best ``glue'' languages for fast prototyping.
Code development in Python is much faster and easier than with many other
languages. Less bugs are introduced by the developers simply because there are
fewer lines of code \citep{nanz_comparative_2015}.
%
Maintaining and modifying existing code is also eased by the very good code
readability associated with the language.
%
Another consequence of the apparent simplicity of Python is that the learning
curve has a nice shape. Learning Python is very easy at the beginning and the
complexity gradually increases as the level of the developer increases.
%
Its interpreted nature is advantageous for a rapid development cycle and the
possibility of interactive workflow (usually with IPython or Jupyter
notebooks).
%
Therefore, Python is good for developers of all levels: very gentle for
beginners and very powerful for advanced users.

Python is an old language (first implementation in 1991!) but it continues to
evolve.
%
Note that there are many implementations of Python
interpreters\footnote{Notable ones are CPython (written in C), Jython (Java),
IronPython (C\#), PyPy (Rpython, a subset of Python) and
MicroPython/CircuitPython (C, targeted to micro-controllers).}. The default and
most widely used implementation is written in C and is called CPython.
%
A hard and controversial decision was taken to clean up Python from its
historical incoherences by having a new series of backwards-incompatible
versions of the language, starting with Python 3.0, released in December 2008.
After a long and difficult process, the transition from Python 2 to Python 3 is
now (201) behind us for nearly all important packages\footnote{Close to
\href{https://python3wos.appspot.com/}{95\% of the most downloaded packages}
are Python 3 compatible now.}. Especially for scientific applications, we can
now work only in Python 3. For instance, \href{https://github.com/%
numpy/numpy/blob/master/doc/neps/dropping-python2.7-proposal.rst}{future
releases of \Numpy} and of \href{http://www.python3statement.org/}{the main
scientific packages} will soon drop compatibility with Python 2. These changes
open doors to very clean, coherent and potentially faster Python
experience. Therefore, one should not use Python 2 for science any more and
instead, adopt the newer versions.
%
New versions of Python have nice new features to adapt to new usages and to
recent trends in computer science.  For example, three new features were
introduced in Python 3.5 (first released on September 2015): the \codeinline{@}
operator for matrix multiplication, the new \codeinline{async} and
\codeinline{await} keywords for concurrency and type hinting\footnote{See also our
paragraph on the issue of type checking in the following.} (with the module
\codeinline{typing} and an associated syntax presented in the
\href{https://www.python.org/dev/peps/pep-0484/}{PEP~484}).

@@

Python
% has recently become one of the most popular language for teaching programming
% and developing scientific applications. It 
has now a mature and powerful scientific ecosystem with well-established based
packages (\Numpy for N-dimensional homogeneous arrays, \Scipy as the
fundamental toolkit for scientific computing, \pack{Matplotlib} for plotting
and \pack{Pandas} for data structures) and several more specialized packages
(to name a few, \pack{h5py}, \pack{mpi4py}, \pack{skimage}, \pack{sklearn}).
The \href{https://github.com/rougier/python-visualization-landscape}{%
landscape of visualization tools usable through Python} makes it possible to
integrate post-processing and analysis capabilities into a single application.
%
Python is one of the main languages for data science with packages such as
\pack{Pandas}, \pack{statmodels}, \pack{sklearn}, \pack{Keras},
\pack{TensorFlow} and \pack{PyTorch}.
%
There user-friendly and ready-to-use Python installers (similar to Matlab), in
particular the
\href{https://en.wikipedia.org/wiki/Anaconda_(Python_distribution)}{%
open-source distribution Anaconda} to streamline the installation in 
desktop platforms.
Integrated development environments (IDE) adapted for
scientists, for example \href{https://github.com/spyder-ide/spyder}{Spyder},
\href{https://pyzo.org}{Pyzo},
and \href{https://jupyterlab.readthedocs.io}{JupyterLab} are often useful for
interactive development.

Python has a very large and supportive community (see for example the
\href{http://stackoverflow.com/tags}{Stack Overflow tags}).
%
But Python development is also supported by companies specialized in web
technologies and data-driven research. However, recently, companies involved in
scientific software development have started to become quite influential. For
example Anaconda Inc.\ is strongly engaged in the development of tools like JIT
Python accelerator \pack{Numba}, and parallelism framework \pack{dask}.
Hardware giants such as Intel and Nvidia also contribute to the Python
scientific ecosystem.
%
To summarize, it is practical to use and extend Python for scientific purposes.

@@

\item \textbf{Performance in CPU bounded tasks.} Since the standard Python
	interpreter does not do any proper compilation, pure Python code for
	some CPU bounded tasks can be too slow. Thus, one has to avoid writing
	loops as much as possible and use vectorized array operations as much
	as possible, or call \emph{foreign functions}, or use special tools to
	build extensions using \emph{Ahead Of Time} or \emph{Just In Time}
	compilation to speedup performance-critical code.

\item \textbf{Absence of type checking.} In contrast to compiled languages,
	there is no type-checking in Python (and more generally in many
	interpreted languages). Types are respected however, and type errors,
	if any, would occur during runtime.
	\emph{Type hints} and \emph{static type checking} can solve this issue.

@@

\subsubsection{Ahead Of Time compilation}
Ahead Of Time compilation is process of generating compiled extensions which
can be dynamically executed by Python.
%
Python developers do not actually have to write the extensions in C, C++ or
Fortran, since there are tools to generate them automatically from Python code
(Pythran) or from Cython code. Cython \citep{behnel_cython2011} is a smooth
blend of two languages with possibility to add type declarations like in C/C++,
but with a syntax similar to Python.
%
There are also other projects to create extensions in C++ with
\href{https://github.com/pybind/pybind11/}{pybind11}, and in Rust with
\href{https://github.com/PyO3}{PyO3}.

Pythran \citep{guelton2015pythran, guelton2018pythran} is a quite recent Python
compiler which gives impressive results. It creates compiled extensions from pure
Python code with simple type annotations written as comments.
%
The resulting extensions are usually as fast as Fortran or C++ written by
non-specialists \citep[see, for example, benchmarks for the packages
\fluidpack{fft} and \fluidpack{sim},][]{fluidfft, fluidsim}.
%
They are created with a very interesting two-step compilations: first the code
is optimized at the Python level and then an automatically produced C++ code is
properly compiled.
%
This two-step compilation provides room for very clever optimizations
\footnote{As demonstrated in this blog post by the main developer of
Pythran \url{http://serge-sans-paille.github.io/pythran-stories/%
being-more-than-a-translator.html}}.
%
Pythran supports OpenMP pragma \citep{guelton2013compiling} and can use modern
vectorization with SIMD instructions \citep{guelton2014exploring}. Pythran
understands both Matlab-like vectorized code and C-like code with explicit loops
\citep{guelton2018pythran}.
%
We present in \citet{fluidfft, fluidsim} examples of
highly efficient Python codes, optimized using a combination of Cython and
Pythran extensions.

\subsubsection{Just In Time (JIT) compilation}
Another way to obtain optimized machine instructions is by using JIT
compilation, i.e.\ by compiling only the critical code at run time.
%
This strategy can yield good results with other languages as for example Matlab
or Julia.
%
PyPy, an alternative interpreter written in Python, has a JIT compiler. However,
PyPy is not widely used for scientific applications mainly because of
compatibility problems with the extensions written for CPython.
%
However this could change since recent versions, starting with PyPy 6.0,
supports the main packages of the Python scientific stack (\Numpy, \Scipy,
\pack{Matplotlib}, \pack{Pandas}, etc.).

Adding a JIT to the interpreter CPython has been notoriously difficult
\footnote{\url{https://faster-cpython.readthedocs.io/}}. One of the
problem seems to be the C API provided in CPython and used by many extensions.
%
A faster CPython with a JIT is actually not so necessary for many purposes,
since Python extensions are enough. Therefore, not as much money and work
has been put in accelerating Python than for example for Java and JavaScript.

Another strategy is to add a JIT to CPython through an external package
(\pack{Numba}).
%
\pack{Numba} is particularly interesting because it can take advantage of the
GPU\footnote{See for example
\url{https://devblogs.nvidia.com/parallelforall/seven-things-numba/}}.

\subsubsection{Type hints and static type checking}
%
Type-checking is useful since it can avoid simple bugs.
\fnref{https://docs.python.org/3/whatsnew/3.5.html\#whatsnew-pep-484}{PEP-484}
introduced a syntax and a related \codeinline{typing} module to add
type-hints, typically for function parameters and return types.
%
Static type checkers like \href{http://mypy-lang.org/}{Mypy} and
\href{https://github.com/google/pytype}{pytype}
%
\footnote{We plan to add type hinting in the most important FluidDyn modules to
investigate how we can take advantage of type-checking with Mypy.}
%
allows to process type-hints and report inconsistencies in the code.

@@

\subsubsection{Achieving computational parallelism}

At any rate, threads using the interpreter do not use the CPU at the same time
so it is not possible to use threads to do multi-core computational parallelism
with pure python code to accelerate CPU bounded tasks.
%
Note however, that threads can be (and are widely) used in Python for
concurrency, i.e.\ to perform I/O tasks (which does not need CPU) concurrently.
%
To do proper computational parallelism for CPU bounded tasks with the GIL, one
has to use other strategies: extensions (fine grain parallelism),
multiprocessing (coarse grain parallelism) and inter-process communication, for
example, with MPI (with mpi4py) or ZeroMQ.

@@

Fortran remains widely used for computing codes, at least in fluid mechanics.
C++ is used in many recent developments. Its recent improvements (C++11, C++14)
make it a modern and very useful language.
%
JavaScript, Java, Scala, Smalltalk, Haskell, R, Julia, Perl and Lua are used for
some scientific codes. Go and Rust are quite young languages with emphasis on
memory-safety and are not (yet) widely used for scientific applications.
%
It is, of course, very difficult to predict which technologies will be widely
used in even the near future (10 or 20 years for example).
%
It seems that the idea of ``one language to do everything for science'' will
not succeed, at least not in the near future. 
%
The limited scope of the language is likely the reason why the community
around Fortran and Julia are not widespread outside academia.
%
It seems that a very important aspect in the future will be interoperability
between coexisting tools\footnote{See for example the cross-language development
platform for in-memory data, \href{https://arrow.apache.org/}{Apache Arrow}.}.
%
Knowledge of Python is also an advantage in terms of employability these days.
%
Considering what we have presented on Python, we think that it is reasonable to
bet that it will become one of the \textit{lingua franca} in science and in
fluid dynamics.
%
We have, therefore, chosen it to be the main language for the project FluidDyn.

@@

We study fluid mechanics via laboratory experiments, in situ measurements,
analytic tools, numerical simulations and data processing.
%
Nowadays, programming is involved in all methods, and thus it is difficult to do
fluid mechanics without software development.
%
However, the level in software engineering is on average very low in the
community.
%
Even today, it is quite commonplace to start a Ph.D. without any serious
training in GNU/Linux and modern programming tools.
%
A large majority of scientists and technicians are also unaware of the new
challenges and opportunities of open-source.

@@

Despite the reliance on commercial software for research in fluid mechanics,
%
there are also a lot of codes produced by scientists in fluid mechanics. But
without the technical knowledge on how to work collectively, such codes
are often of bad quality (compared to the information technology
industry standard) and are basically doomed to be abandoned after their use.
%
However, it is natural to try to reuse codes or at least to build the next steps
on what has been done already. Thousands of hours of highly qualified people
are spent in trying to understand and reuse badly written codes in
inappropriate languages!

From a technical perspective, the majority of the coding in the field involves
a mix of Fortran/C or C++, shell languages (as Bash) and Matlab. For
experiments, the graphical programming environment
\href{http://www.ni.com/en-us/shop/labview.html}{Labview} is dominant for
control of physical objects and acquisition. Fortran is often used to implement
solvers and post-processing algorithms. Matlab is used for data processing and
visualization.
%
Often, languages are used for things for which they are not adapted.
%
Fortran, C or C++ have not been designed for fast prototyping of complex
programs.
%
It is known that for scientific purposes, one should restrict shell scripting
to extremely simple tasks.
%
Use of compiled languages to develop new algorithms hinders the development
process and ability to debug.
%
Similarly, using Matlab for developing complex programs is a questionable
choice.
%
These technical solutions are also problematic for code reuse, sharing and
collaborative development.
%
We are now going to present factual arguments explaining why we think that the
massive usage of Matlab decreases the collective efficiency of the community.

@@

FluidDyn is a project to foster open-science and open-source coding in Python
in the field of fluid mechanics.
%
The project envisages to provide the technical framework to allow collaborative
development of tools useful for the fluid mechanics community, and to do better
science with open methods.
%
What have not been ``reinventing the wheel'', but rather have been ``standing
on the shoulders of giants''. In other words we have utilized and extended
existing and mature packages and libraries to create packages in FluidDyn as
demonstrated in figure \ref{fig:dependency}. We have designed them such that
the packages are generic enough for a wider audience and are maintainable in
the long run.

>>>>

Collaboration was achieved in the early years of FOSS development through
emailing ``patches'', and centralized Version Control Systems (VCS) near the
turn of the 21\textsuperscript{st} century.  The collective collaboration on
development is today streamlined with the emergence of Distributed Version
Control Systems (DVCS), especially, Git and Mercurial).
%
Version control tools are designed to work on any kind of text files, not just
codes, so they could even be utilized to write scientific documents with
multiple authors.
%
In web-based source development platforms like GitHub, Bitbucket and GitLab,
DVCS repositories are enhanced with code-review tools, typically consisting of
a issue tracker and pull-request mechanism.
%
Git has become the \textit{de facto} standard for source management. However,
for the FluidDyn project, we prefer to use Mercurial because we consider it to
be simpler for the beginners and just as powerful for expert users\footnote{See
  \href{https://www.mercurial-scm.org/wiki/GitConcepts}{``a detailed
comparison''} on Mercurial website.}. Simplicity and lowering the learning
barrier for scientists who are not experts in software development is crucial
for our project.
%
Note that, it is now possible to use Mercurial to work with Git repositories,
for example hosted in GitHub and GitLab\footnote{By using the Mercurial
extension \href{http://hg-git.github.io/}{hg-git}.}.
%
The use of DVCS in science will lead to long term productivity benefits
\citep{wilson_best_2014}.

@@

Packaging process depends on the language used in a project, however some good
practices are universal, such as licensing, supplying documentation (a README
file at the bare minimum), keeping installation paths flexible and relying on
official build tools \citep{taschuk_ten_2017}.
%
A credible repository to packages as compressed archives or as pre-compiled binary
packages is indispensable to promote code reuse and reproducibility of results.
%
The use of repositories started with the conception of
CPAN and CRAN repositories for the languages Perl and R in 1993 --- based on
the Comprehensive \TeX\ Archive Network (CTAN) model for \TeX\ packages. 
%
Python as a medium for scientific computing owes its success to a rich
ecosystem of third-party packages. \href{https://pypi.org}{Python Package
Index (PyPI)} and \href{https://anaconda.org}{Anaconda Cloud} are major
repositories for delivering and downloading Python packages, assisted by
package managers \pack{pip} and \pack{conda} respectively. The
\fnref{https://packaging.python.org}{``Python Packaging User Guide''} is an
excellent guide to learn how to install, package and distribute projects.
% Almost all major
% languages\footnote{\href{https://stackoverflow.com/questions/1693529/%
% list-of-top-repositories-by-programming-language}{https://stackoverflow.com/%
% questions/1693529/list-of-top-repositories-by-programming-language}} have
% similar repositories, with exceptions such as Fortran and Matlab.
Specialized package managers, such as \pack{easybuild} and \pack{spack}, have
also emerged to install scientific libraries and compilers.

\subsubsection{Continuous Integration and documentation} The need for
strong reliability of packages with increasing complexity has forced developers
to use Continuous Integration (CI).
%
When the code depends on a wide variety of external dependencies, CI is useful
to avoid \fnref{https://en.wikipedia.org/wiki/Software\_rot}{software rot},
i.e.\ to avoid obsolescence with newer versions of the compiler or its
dependencies. CI is specifically important for dynamical languages and
open-source code since, it allows open-source projects to welcome contributions
from other developers while maintaining a high degree of reliability.
%
Firstly the code is \emph{built} against a known, repeatable emulation
environment on a server. Thereafter, a set of \emph{unit tests} are run for
each commit (modification of the code) to check that no bugs have been
introduced. It is also important to analyse what portion of the source code is
covered by these unit tests --- a method referred to as \emph{code coverage}.
%
Almost all CI platforms are free for open-source projects --- Travis CI,
CircleCI and Appveyor to name a few. Bitbucket and GitLab have built-in CI
implementations. Websites such as Codecov and Coveralls help to chart and
analyse the code coverage history, viewable by anyone. To use these services,
one simply has to write a simple file in YAML format and authenticate the
respective services with ``read'' permissions.

@@

% Mutual aid and knowledge sharing have been important aspects of the web since
% its inception and these practices are now of foremost importance for many
% human activities.
%
%In particular, 
Learning programming and computer science is now greatly aided by websites like
the community driven encyclopaedia \href{https://www.wikipedia.org/}{Wikipedia}
or the community driven forum \href{https://stackoverflow.com/}{Stack
Overflow}, where people write thousands of questions and answers on programming
each week. IRC protocol has continued to sustain the test of time. Instant
messaging channels such as \codeinline{\#python} on Freenode facilitate
interactive discussions and a form of community-level support system --- a role
which used to be filled by mailing lists alone in the past. Alternatives such
as Riot (Matrix protocol), Gitter, and Slack have gained popularity offering
different integration solutions on top of instant messaging.
%
Through instant messaging a geographically separated team, as in the case of
the FluidDyn project, can achieve near real-time collaboration without having
to arrange video conferences or meetings.

@@

\item \textbf{Code style.} There are official (but, optional) guidelines, named
\fnref{https://www.python.org/dev/peps/pep-0008/}{PEP~8}, advocating for
regularity of the code. Linters and code-formatters can evaluate and enforce
adherence to this style.

@@

\item \textbf{Extensible}. It is easy to interface with code written in
several other languages (in particular C, C++ and Fortran).

@@

\end{itemize}
%
Thus, Python is one of the best ``glue'' languages for fast prototyping.
Code development in Python is much faster and easier than with many other
languages. Less bugs are introduced by the developers simply because there are
fewer lines of code \citep{nanz_comparative_2015}.
%
Maintaining and modifying existing code is also facilitated by the very good code
readability associated with the language.
%
Another consequence of the apparent simplicity of Python is that the learning
curve has a nice shape. Learning Python is very easy from the start and the
complexity gradually increases as the level of the developer increases.
%
Its interpreted nature is advantageous for a rapid development cycle and the
possibility of interactive workflow (usually with IPython or Jupyter
notebooks).
%
Therefore, Python is good for developers of all levels: very gentle for
beginners and very powerful for advanced users.

Python is an old language (first implementation in 1991!) but it continues to
evolve.
%
Note that there are many implementations of Python
interpreters\footnote{Notable ones are CPython (written in C), Jython (Java),
IronPython (C\#), PyPy (Rpython, a subset of Python) and
MicroPython/CircuitPython (C, targeted to micro-controllers).}. The default and
most widely used implementation is written in C and is called CPython.
%
A hard and controversial decision was taken to clean up Python from its
historical incoherences by having a new series of backwards-incompatible
versions of the language, starting with Python 3.0, released in December 2008.
After a long and difficult process, the transition from Python 2 to Python 3 is
behind us for nearly all important packages\footnote{Close to
\href{https://python3wos.appspot.com/}{95\% of the most downloaded packages}
are Python 3 compatible as of 2018.}. Especially for scientific applications,
we can now work only in Python 3. For instance, \href{https://github.com/%
numpy/numpy/blob/master/doc/neps/dropping-python2.7-proposal.rst}{future
releases of \Numpy} and of \href{http://www.python3statement.org/}{the main
scientific packages} will soon drop compatibility with Python 2. These changes
open doors to very clean, coherent and potentially faster Python
experience. Therefore, one should not use Python 2 for science any more and
instead, adopt the newer versions.
%
New versions of Python have nice new features to adapt to new usages and to
recent trends in computer science.  For example, three new features were
introduced in Python 3.5 (first released on September 2015): the \codeinline{@}
operator for matrix multiplication, the new \codeinline{async} and
\codeinline{await} keywords for concurrency and type hinting\footnote{See also our
paragraph on the issue of type checking in the following.} (with the module
\codeinline{typing} and an associated syntax presented in the
\href{https://www.python.org/dev/peps/pep-0484/}{PEP~484}).

@@

\end{itemize}
%
Python
% has recently become one of the most popular language for teaching programming
% and developing scientific applications. It 
has now a mature and powerful scientific ecosystem with well-established
foundational packages
%
(\Numpy for N-dimensional homogeneous arrays, \Scipy as the
fundamental toolkit for scientific computing, \pack{Matplotlib} for plotting
and \pack{Pandas} for data structures) and several more specialized packages
(to name a few, \pack{h5py}, \pack{mpi4py}, \pack{skimage}, \pack{sklearn}).
The \href{https://github.com/rougier/python-visualization-landscape}{%
landscape of visualization tools usable through Python} makes it possible to
integrate post-processing and analysis capabilities into a single application.
%
Python is one of the main languages for data science with packages such as
\pack{Pandas}, \pack{statmodels}, \pack{sklearn}, \pack{Keras},
\pack{TensorFlow} and \pack{PyTorch}.
%
There user-friendly and ready-to-use Python installers (similar to Matlab), in
particular the
\href{https://en.wikipedia.org/wiki/Anaconda_(Python_distribution)}{%
open-source distribution Anaconda} to streamline the installation in 
desktop platforms.
Integrated development environments (IDE) adapted for
scientists, for example \href{https://github.com/spyder-ide/spyder}{Spyder},
\href{https://pyzo.org}{Pyzo},
and \href{https://jupyterlab.readthedocs.io}{JupyterLab} are often useful for
interactive development.

Python has a very large and supportive community (see, for example,
\fnref{http://stackoverflow.com/tags/python}{Stack Overflow tags}).
%
Python development is also supported by companies, and most of them specialize
in web technologies and data-driven research. However, recently, companies
involved in scientific software development have started to become quite
influential. For example, Anaconda Inc.\ is strongly engaged in the development
of tools like JIT Python accelerator \pack{Numba}, and parallelism framework
\pack{dask}. Hardware giants such as Intel and Nvidia also contribute to the
Python scientific ecosystem.
%
To summarize, it is practical to use and extend Python for scientific purposes.

@@

\item \textbf{Performance in CPU bounded tasks.} Since the standard Python
	interpreter does not do any proper compilation, pure Python code for
	some CPU bounded tasks can be too slow. Thus, one has to avoid writing
	loops as much as possible and use vectorized array operations as much
	as possible. One can also call \emph{foreign functions}, or use special
	tools to build extensions using \emph{Ahead Of Time} or \emph{Just In
	Time} compilation to speedup performance-critical code.

\item \textbf{Absence of type checking.} In contrast to compiled languages,
	there is no type-checking in Python (and more generally in many
	interpreted languages). Types are respected however, and type errors,
	if any, would be raised during runtime.
	\emph{Type hints} and \emph{static type checking} can solve this issue.

@@

\subsubsection{Ahead Of Time compilation}
Ahead Of Time compilation is process of generating compiled extensions which
can be dynamically executed by Python.
%
Python developers do not actually have to write the extensions in C, C++ or
Fortran, since there are tools to generate them automatically from Python code
(Pythran) or from Cython code. Cython \citep{behnel_cython2011} is a smooth
blend of two languages with possibility to add type declarations like in C/C++,
but with a syntax similar to Python.
%
There are also other projects to create extensions in C++ with \pack{pybind11},
and in Rust with \href{https:///pyo3.rs}{PyO3}.

Pythran \citep{guelton2015pythran, guelton2018pythran} is a quite recent Python
compiler which gives impressive results. It creates compiled extensions from pure
Python code with simple type annotations written as comments.
%
The resulting extensions are usually as fast as Fortran or C++ written by
non-specialists \citep[see, for example, benchmarks for the packages
\fluidpack{fft} and \fluidpack{sim},][]{fluidfft, fluidsim}.
%
They are created with a very interesting two-step compilations: first the code
is optimized at the Python level and then an automatically produced C++ code is
properly compiled.
%
This two-step compilation provides room for very clever optimizations%
\footnote{As demonstrated in this blog post by the main developer of
Pythran \url{http://serge-sans-paille.github.io/pythran-stories/%
being-more-than-a-translator.html}}.
%
Pythran supports OpenMP pragma \citep{guelton2013compiling} and can use modern
vectorization with SIMD instructions \citep{guelton2014exploring}. Pythran
understands both Matlab-like vectorized code and C-like code with explicit loops
\citep{guelton2018pythran}.
%
We present in \citet{fluidfft, fluidsim} examples of
highly efficient Python codes, optimized using a combination of Cython and
Pythran extensions.

\subsubsection{Just In Time (JIT) compilation}
Another way to obtain optimized machine instructions is by using JIT
compilation, i.e.\ by compiling only the critical code at run time.
%
This strategy can yield good results with other languages as for example Matlab
or Julia.
%
Adding a JIT to the interpreter CPython has been notoriously difficult
\footnote{\url{https://faster-cpython.readthedocs.io/}}. One of the
problem seems to be the C API provided in CPython and used by many extensions.
%
A faster CPython with a JIT is actually not so necessary for many purposes,
since Python extensions are enough. Therefore, not as much money and work
has been put in accelerating Python than for example for Java and JavaScript.

PyPy, an alternative interpreter written in Python, has a JIT compiler. However,
PyPy is not widely used for scientific applications mainly because of
compatibility problems with the extensions written for CPython.
%
However this could change since recent versions, starting with PyPy 6.0,
supports the main packages of the Python scientific stack (\Numpy, \Scipy,
\pack{Matplotlib}, \pack{Pandas}, etc.).
%
Another strategy is to add a JIT to CPython through an external package
(\pack{Numba}).
%
\pack{Numba} is particularly interesting because it can take advantage of the
GPU\footnote{See for example
\url{https://devblogs.nvidia.com/parallelforall/seven-things-numba/}}.

\subsubsection{Type hints and static type checking}
%
Type-checking is useful since it can avoid simple bugs.
\fnref{https://docs.python.org/3/whatsnew/3.5.html\#whatsnew-pep-484}{PEP-484}
introduced a syntax and a related \codeinline{typing} module to add
type-hints, typically for function parameters and return types.
%
Static type checkers like \href{http://mypy-lang.org/}{Mypy} and
\href{https://github.com/google/pytype}{pytype}%
\footnote{We plan to add type hinting in the most important FluidDyn modules to
investigate how we can take advantage of type-checking with Mypy.}
%
allows to process type-hints and report inconsistencies in the code.

@@

\subsubsection{Multi-core computational parallelism}

At any rate, threads using the interpreter do not use the CPU at the same time
so it is not possible to use threads to do multi-core computational parallelism
with pure python code to accelerate CPU bounded tasks.
%
Note however, that threads can be (and are widely) used in Python for
concurrency, i.e.\ to perform I/O tasks, which does not need CPU, concurrently.
%
To do proper computational parallelism for CPU bounded tasks with the GIL, one
has to use other strategies: extensions (fine grain parallelism),
multiprocessing (coarse grain parallelism) and inter-process communication, for
example, with MPI (with mpi4py) or ZeroMQ.

@@

Fortran remains widely used for computing codes, at least in fluid mechanics.
C++ is used in many recent developments. Its recent improvements (C++11, C++14)
make it a modern and very useful language.
%
JavaScript, Java, Scala, Smalltalk, Haskell, R, Julia, Perl and Lua are used for
some scientific codes. Go and Rust are quite young languages with emphasis on
memory-safety and are not (yet) widely used for scientific applications.
%
It is, of course, very difficult to predict which technologies will be widely
used in even the near future (10 or 20 years for example).
%
It seems that the idea of ``one language to do everything for science'' will
not succeed, at least not in the near future. 
%
The limited scope of the language is likely the reason why the community
around Fortran and Julia are not widespread outside academia.
%
An important aspect in the future would very likely be interoperability
between coexisting tools\footnote{See for example the cross-language development
platform for in-memory data, \href{https://arrow.apache.org/}{Apache Arrow}.}.
%
Considering what we have presented on Python, we think that it is reasonable to
bet that it will become one of the \textit{lingua franca} in science and in
fluid dynamics.
%
We have, therefore, chosen it to be the main language for the project FluidDyn.

@@

We study fluid mechanics via laboratory experiments, in situ measurements,
analytic tools, numerical simulations and data processing.
%
Nowadays, programming is involved in all methods, and is usually difficult to do
fluid mechanics without software development.
%
However, the level in software engineering is on average very low in the
community.
%
Even today, it is quite commonplace to start a Ph.D. without any serious
training in GNU/Linux and modern programming tools.
%
A large majority of scientists and technicians are also unaware of the new
challenges and opportunities of open-source.

@@

Despite the reliance on commercial software for research in fluid mechanics,
%
there are also a lot of codes produced by scientists in fluid mechanics. But
without the technical knowledge on how to work collectively, such codes
are often substandard (compared to the information technology
industry standards) and are basically doomed to be abandoned after their use.
%
However, it is natural to try to reuse codes or at least to build the next steps
on what has been done already. Thousands of hours of highly qualified people
are spent in trying to understand and reuse badly written codes in
inappropriate languages!

From a technical perspective, the majority of the coding in the field involves
a mix of Fortran/C or C++, shell languages (as Bash) and Matlab. For
experiments, the graphical programming environment
\href{http://www.ni.com/en-us/shop/labview.html}{Labview} is dominant for
control of physical objects and acquisition. Fortran is often used to implement
solvers and post-processing algorithms. Matlab is used for data processing and
visualization.
%
Often, languages are used for purposes which they are not adapted for.
%
Fortran, C or C++ have not been designed for fast prototyping of complex
programs.
%
It is known that for scientific purposes, one should restrict shell scripting
to extremely simple tasks.
%
Use of compiled languages to develop new algorithms hinders the development
process and ability to debug.
%
Similarly, using Matlab for developing complex programs is a questionable
choice.
%
These technical solutions are also problematic for code reuse, sharing and
collaborative development.
%
We are now going to present factual arguments explaining why we think that the
massive usage of Matlab decreases the collective efficiency of the community.

@@

FluidDyn is a project to foster open-science and open-source coding in Python
in the field of fluid mechanics.
%
The project envisages to provide the technical framework to allow collaborative
development of tools useful for the fluid mechanics community, and to do better
science with open methods.
%
We have not been ``reinventing the wheel'', but rather have been ``standing
on the shoulders of giants''. In other words, we have utilized and extended
existing and mature packages and libraries to create packages in FluidDyn as
demonstrated in figure \ref{fig:dependency}. We have designed them such that
the packages are generic enough for a wider audience and are maintainable in
the long run.
